---
title: "Real-time CFR over time"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Real-time CFR over time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The case fatality ratio often changes during an outbreak, as the population
affected, the pathogen or the care available changes. This vignette looks at two
estimates of the CFR for each week of onset. The first is `prob`, the
probability of death the fitted model gives a case from that week; `fit_cfr()`
estimates it, with a formula to let it vary over time. The second is the
fraction of that week's cases that will die, given what has been observed about
them so far. From each case's probability of death, which
`posterior_prob_death()` gives, we get a predicted fatal fraction for each week.

The two answer different questions and draw on different information. We
compare them on a simulated outbreak whose CFR rises late, fitted at two
cut-offs.

## Simulation setup

Onsets are spread over twelve weeks. The true CFR is 0.25 up to week 7 and 0.5
from week 8. The onset-to-death delay is lognormal with a mean of about 13 days.

```{r simulate, eval = FALSE}
library(cfrnow)
set.seed(20260930)

n <- 2000
onset_start <- as.Date("2026-01-01")
true_cfr <- function(week) ifelse(week >= 8, 0.5, 0.25)

onset_day <- sample.int(12 * 7, n, replace = TRUE) - 1
onset_date <- onset_start + onset_day
week <- onset_day %/% 7
fatal <- runif(n) < true_cfr(week)

death_date <- as.Date(rep(NA, n))
death_date[fatal] <- onset_date[fatal] +
  floor(runif(sum(fatal)) + rlnorm(sum(fatal), 2.41, 0.51))
ll <- data.frame(onset_date = onset_date, death_date = death_date)
```

## Two fits at a cut-off

We prepare the data one week after the last onset, add the onset week as a
column, and fit two models: one with a constant `prob`, and one where `prob`
follows a natural spline in the onset week.

```{r fit, eval = FALSE}
cure <- prepare_cfr_data(ll, obs_time = max(ll$onset_date) + 7) |>
  as_epidist_cure_model()
cure$week <- as.numeric(cure$onset - onset_start) %/% 7

onset_to_death <- LogNormal(
  meanlog = Normal(2.4, 0.2), sdlog = Normal(0.5, 0.15)
)
fit_constant <- fit_cfr(cure,
  delay = onset_to_death,
  formula = brms::bf(mu ~ 1, prob ~ 1)
)
fit_spline <- fit_cfr(cure,
  delay = onset_to_death,
  formula = brms::bf(mu ~ 1, prob ~ splines::ns(week, df = 3))
)
```

## From per-case probabilities to a weekly CFR

For a fit, `brms::posterior_epred(fit, dpar = "prob")` gives the draws of `prob`
for every case. `posterior_prob_death(fit)` gives the draws of each case's
probability of death. A death already recorded counts as 1 and a recovery as 0.
A case still unresolved at the cut-off gets its `prob` updated by the time
passed since symptom onset: the longer ago symptoms began, the lower its
probability. Averaging these probabilities over a week's cases gives the fatal
fraction the model expects for that week. For an interval on the fraction that
actually dies, we also draw each case's outcome from its probability, once per
posterior draw, and take the fraction of fatal outcomes in each week. We call
this the predicted fatal fraction. Its median is close to the expected fraction,
but its interval also covers whether each unresolved case dies.

```{r nowcast, eval = FALSE}
weekly <- function(draws, week) {
  sapply(split(seq_along(week), week), function(i) {
    rowMeans(draws[, i, drop = FALSE])
  })
}
week_of_case <- as.numeric(fit_spline$cfrnow$onset - onset_start) %/% 7

pi <- posterior::as_draws_matrix(posterior_prob_death(fit_spline))
fatal <- matrix(rbinom(length(pi), 1, pi), nrow(pi))
nowcast <- weekly(fatal, week_of_case)
prob <- weekly(brms::posterior_epred(fit_spline, dpar = "prob"), week_of_case)
apply(nowcast, 2, quantile, probs = c(0.05, 0.5, 0.95))
```

The fits and summaries shown below were run by `data-raw/real_time_cfr.R` in
the package sources, which also refits both models six weeks after the last
onset. The vignette reads its saved results.

```{r read, message = FALSE}
library(ggplot2)
rtc <- readRDS(system.file("vignette-data", "real_time_cfr.rds",
  package = "cfrnow"
))

plot_weekly <- function(cut) {
  est <- rtc$estimates[rtc$estimates$cutoff == cut, ]
  est$quantity <- factor(est$quantity,
    levels = c("prob", "nowcast"),
    labels = c("prob", "predicted fatal fraction")
  )
  naive <- rtc$naive[rtc$naive$cutoff == cut, ]
  ggplot(est, aes(week)) +
    geom_ribbon(aes(ymin = lower, ymax = upper, fill = quantity), alpha = 0.3) +
    geom_line(aes(y = median, colour = quantity), linewidth = 0.8) +
    geom_step(
      data = rtc$truth, aes(x = week - 0.5, y = true_cfr), direction = "hv",
      linetype = "dashed"
    ) +
    geom_point(data = rtc$truth, aes(y = realised), shape = 4, size = 2) +
    geom_point(data = naive, aes(y = naive), colour = "grey50") +
    facet_wrap(~model) +
    scale_colour_brewer(palette = "Dark2") +
    scale_fill_brewer(palette = "Dark2") +
    labs(
      x = "onset week", y = "CFR", colour = NULL, fill = NULL,
      caption = paste(
        "Dashed: true CFR. Crosses: fraction of each week's cases that die.",
        "Grey points: deaths / cases at the cut-off."
      )
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")
}
```

## One week after the last onset

```{r early, fig.width = 7, fig.height = 4.5, fig.alt = "Weekly CFR estimates one week after the last onset. The constant model's prob is flat at 0.32; the spline's prob dips slightly, then rises from week 5. The predicted fatal fraction matches the realised fractions for the earlier weeks and, for the last two weeks, is pulled towards each model's prob."}
plot_weekly("early")
```

The naive ratio (grey) falls away in the last weeks, because most of their
deaths have yet to happen: 0.22 in week 11, against the 0.47 of that week's
cases that die.

For weeks 0 to 7 the predicted fatal fraction is the same under both models and
has almost no uncertainty. Nearly every case from those weeks has died or has
been followed long enough to be a near-certain survivor, which makes the
prediction the realised fraction. That includes the chance variation between
weeks, such as the 0.32 in week 3, where the true CFR is 0.25. Because most
deaths from weeks 8 and 9 have already happened, those weeks already show the
step: under the constant model the prediction is 0.46 and 0.53, against realised
fractions of 0.48 and 0.55. The deaths still to come account for the remaining
gap.

The models differ most in weeks 10 and 11. Their cases are largely unresolved,
which keeps the probabilities for those cases close to their `prob`:

- The constant model's `prob` of 0.32 is an average over the whole outbreak. It
  pulls the prediction down to 0.44 in week 10 and 0.37 in week 11.
- The spline's `prob` dips slightly over the first weeks, then rises from week
  5 to 0.61 in week 11, where the prediction is 0.58. A natural spline with
  three degrees of freedom cannot follow a step: it starts rising before week 8
  and keeps rising after it, overshooting in the last week.

## Six weeks later

```{r late, fig.width = 7, fig.height = 4.5, fig.alt = "Weekly CFR estimates six weeks after the last onset. The predicted fatal fraction equals the realised fractions in every week but the last under both models; the two models' prob still differ."}
plot_weekly("late")
```

Six more weeks of follow-up give nearly every fatal case time to die and make
the survivors near-certain. For both models, the predicted fatal fraction is now
the realised fraction in every week but the last, where one death has yet to be
recorded. The naive ratio has caught up with it. The two estimates of `prob`
still differ: the constant model gives 0.33 in every week, while the spline
reaches 0.53 in week 11 at the end of a now flatter ramp. Each estimates the
underlying CFR, which more follow-up only narrows.

## Interpreting the two estimates

`prob` is the model's CFR for a case from that week. With a constant formula it
is one number for the whole outbreak. With a time-varying formula it is a curve
whose last weeks rest on cases with little follow-up, which makes them depend on
the shape the formula allows.

The predicted fatal fraction estimates the fraction of that week's cases that
die. For a recorded death it uses what happened; for a case still unresolved it
falls back on the model's `prob`, lowered by the time passed since symptom
onset. Its interval covers the model's parameters and whether each unresolved
case dies, but it assumes the model is right: in week 11 at the early cut-off,
the realised 0.47 lies outside the interval of both models. Once the fatal cases
have died and the survivors have been followed well past the delay, the interval
is close to zero width, even though the underlying CFR is still uncertain.

In practice:

- Report the predicted fatal fraction for the cases seen so far, and `prob` for
  the underlying CFR, for instance to compare groups or project to new cases.
- For the most recent weeks the predicted fatal fraction is only as good as the
  model's `prob` for them. With a constant formula, a recent change shows up
  only as its cases resolve. A time-varying formula lets it show earlier, but at
  the cost of depending on the shape the formula allows.
- Refit with more than one formula and see how much the recent weeks move.