---
title: "Real-time CFR over time"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Real-time CFR over time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# Fit chunks need CmdStan; evaluate them only where a toolchain is present so the
# vignette still builds on CmdStan-less builders, where it shows the code
# without running it.
has_cmdstan <- requireNamespace("cmdstanr", quietly = TRUE) &&
  !is.null(tryCatch(cmdstanr::cmdstan_version(), error = function(e) NULL))
knitr::opts_chunk$set(eval = has_cmdstan)
# numbers quoted in the text come from the fits; without them, show a gap
est <- function(cut, model, qty, wk, stat = "median") {
  if (!exists("estimates")) {
    return("...")
  }
  row <- estimates[
    estimates$cutoff == cut & estimates$model_name == model &
      estimates$quantity == qty & estimates$week == wk,
  ]
  sprintf("%.2f", row[[stat]])
}
interval <- function(cut, model, qty, wk) {
  paste0(
    est(cut, model, qty, wk), " [", est(cut, model, qty, wk, "lower"), ", ",
    est(cut, model, qty, wk, "upper"), "]"
  )
}
```

The case fatality ratio often changes during an outbreak, as the population affected, the pathogen or the care available changes.
This vignette looks at two estimates of the CFR for each week of onset:

- `prob`, the probability of death that the fitted model gives a case from that week, estimated by `fit_cfr()`;
- the predicted fatal fraction, the fraction of that week's cases expected to die given what has been observed about them so far, built from `posterior_prob_death()`.

We compare them on a simulated outbreak whose CFR rises late.
One model has a constant `prob`; in the other, `prob` follows a random walk over onset weeks.
We fit both one week after the last onset, then refit the random walk six weeks after it.
The simulation is the simplest setting: deaths only, no recorded recoveries, no loss to follow-up and one cut-off for all cases.
"Other settings" below covers what changes otherwise.

## Simulation setup

Onsets are spread over twelve weeks.
The true CFR is 0.25 up to week 7 and 0.5 from week 8.
The onset-to-death delay is lognormal with a mean of about 13 days.

```{r simulate, eval = TRUE, message = FALSE}
library(cfrnow)
library(data.table)
library(ggplot2)
set.seed(20260930)

n <- 1000
onset_start <- as.Date("2026-01-01")
true_cfr <- function(week) ifelse(week >= 8, 0.5, 0.25)

onset_day <- sample.int(12 * 7, n, replace = TRUE) - 1
onset_date <- onset_start + onset_day
week <- onset_day %/% 7
fatal <- runif(n) < true_cfr(week)

death_date <- as.Date(rep(NA, n))
death_date[fatal] <- onset_date[fatal] +
  floor(runif(sum(fatal)) + rlnorm(sum(fatal), 2.41, 0.51))
ll <- data.frame(onset_date = onset_date, death_date = death_date)

# the true CFR and the fraction of each week's cases that die
truth <- data.table(week = week, fatal = fatal)[
  , .(realised = mean(fatal)),
  keyby = week
][, true_cfr := true_cfr(week)]
```

## Fitting a constant and a random-walk model

`prepare_cfr_data()` classifies each case at a cut-off, here one week after the last onset.
`as_epidist_cure_model()` turns the result into one row per case, to which we add the onset week as a factor.
In the random-walk model, each week's `prob` (on the logit scale) stays close to the weeks either side of it, with the data deciding how close.
brms fits this as a Markov random field smooth (`bs = "mrf"`) whose neighbour list links each week to the next.
A week with few resolved cases follows its neighbours with a wide interval, because its own cases say little yet.

```{r fit-early, results = "hide", message = FALSE, warning = FALSE}
prepare <- function(cut) {
  cure <- prepare_cfr_data(ll, obs_time = max(ll$onset_date) + cut) |>
    as_epidist_cure_model()
  cure$week <- as.numeric(cure$onset - onset_start) %/% 7
  cure$week_factor <- factor(cure$week, levels = weeks)
  cure
}
weeks <- 0:11
neighbours <- lapply(weeks, \(w) as.character(intersect(c(w - 1, w + 1), weeks)))
names(neighbours) <- weeks
random_walk <- brms::bf(
  mu ~ 1, prob ~ s(week_factor, bs = "mrf", xt = list(nb = neighbours))
)
onset_to_death <- LogNormal(
  meanlog = Normal(2.4, 0.2), sdlog = Normal(0.5, 0.15)
)
fit <- function(cure, formula) {
  fit_cfr(cure,
    delay = onset_to_death, formula = formula,
    backend = "cmdstanr", chains = 4, cores = 4, refresh = 0, seed = 1
  )
}
early <- prepare(7)
fit_constant <- fit(early, brms::bf(mu ~ 1, prob ~ 1))
fit_random_walk <- fit(early, random_walk)
```

## Weekly estimates from per-case draws

A `fit_cfr()` fit is a brms model, which brms tools work on.
`brms::posterior_epred(fit, dpar = "prob")` gives the draws of `prob` for every case.
`posterior_prob_death()` gives the draws of each case's probability of death.
A recorded death counts as 1.
A case still unresolved at the cut-off has its `prob` updated by the time passed since symptom onset: the longer ago symptoms began, the lower its probability.

For the predicted fatal fraction, we draw each case's outcome from its probability, once per posterior draw, and take the fraction of fatal outcomes in each week.
Its median is close to the average of the probabilities.
Its interval also covers whether each unresolved case dies.
The function below returns both estimates for each onset week, as medians with 90% intervals.

```{r weekly}
weekly_estimates <- function(fit) {
  week <- as.numeric(fit$cfrnow$onset - onset_start) %/% 7
  prob <- brms::posterior_epred(fit, dpar = "prob")
  pi <- posterior::as_draws_matrix(posterior_prob_death(fit))
  fatal <- matrix(rbinom(length(pi), 1, pi), nrow(pi))
  draws <- data.table(
    draw = seq_len(nrow(pi)), week = rep(week, each = nrow(pi)),
    prob = as.vector(prob), nowcast = as.vector(fatal)
  )
  by_week <- draws[, .(prob = mean(prob), nowcast = mean(nowcast)),
    by = .(draw, week)
  ]
  melt(by_week, id.vars = c("draw", "week"), variable.name = "quantity")[
    , .(
      median = median(value), lower = quantile(value, 0.05),
      upper = quantile(value, 0.95)
    ),
    keyby = .(quantity, week)
  ]
}
set.seed(1)
estimates <- rbind(
  data.table(cutoff = "early", model_name = "constant",
    weekly_estimates(fit_constant)),
  data.table(cutoff = "early", model_name = "random walk",
    weekly_estimates(fit_random_walk))
)
```

```{r plot-helper}
plot_weekly <- function(cut, cure) {
  est_cut <- estimates[cutoff == cut]
  est_cut[, quantity := factor(quantity,
    levels = c("prob", "nowcast"),
    labels = c("prob", "predicted fatal fraction")
  )]
  naive <- as.data.table(cure)[, .(naive = mean(outcome == 1)), keyby = week]
  ggplot(est_cut, aes(week)) +
    geom_ribbon(aes(ymin = lower, ymax = upper, fill = quantity), alpha = 0.3) +
    geom_line(aes(y = median, colour = quantity), linewidth = 0.8) +
    # repeat the last week so the step also spans it
    geom_step(
      data = rbind(truth, truth[.N][, week := week + 1]),
      aes(x = week - 0.5, y = true_cfr), direction = "hv", linetype = "dashed"
    ) +
    geom_point(data = truth, aes(y = realised), shape = 4, size = 2) +
    geom_point(data = naive, aes(y = naive), colour = "grey50") +
    facet_wrap(~model_name) +
    scale_colour_brewer(palette = "Dark2") +
    scale_fill_brewer(palette = "Dark2") +
    labs(
      x = "onset week", y = "CFR", colour = NULL, fill = NULL,
      caption = paste(
        "Dashed: true CFR. Crosses: fraction of each week's cases that die.",
        "Grey points: deaths / cases at the cut-off."
      )
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")
}
```

## Estimates one week after the last onset

```{r early, fig.width = 7, fig.height = 4.5, fig.alt = "Weekly CFR estimates one week after the last onset from a constant and a random-walk model. The constant model's prob is flat; the random walk's prob rises after the step, with intervals widening towards the last week. The predicted fatal fraction matches the realised fractions up to week 7 under both models and, for the last two weeks, is pulled towards each model's prob."}
plot_weekly("early", early)
```

The naive ratio (grey) falls away in the last weeks, because most of their deaths have yet to happen: `r if (exists("early")) sprintf("%.2f", mean(early$outcome[early$week == 11] == 1)) else "..."` in week 11, against the `r sprintf("%.2f", truth[week == 11, realised])` of that week's cases that die.

Up to week 7, the predicted fatal fraction is the same under both models and has almost no uncertainty.
Nearly every case from those weeks has died or has been followed long enough to be a near-certain survivor.
The prediction is then the realised fraction, including its chance variation around the true CFR.
Weeks 8 and 9 still have some unresolved cases, which gives their predictions a narrow interval.

The models differ most in weeks 10 and 11.
Their cases are largely unresolved, which keeps the probabilities for those cases close to their `prob`:

- The constant model's `prob` of `r est("early", "constant", "prob", 11)` is an average over the whole outbreak, before and after the step.
  It pulls the prediction for week 11 down to `r interval("early", "constant", "nowcast", 11)`.
- The random walk's `prob` for week 11 is `r interval("early", "random walk", "prob", 11)`.
  With few resolved cases in that week, it stays close to week 10, with a wide interval.
  The prediction for week 11 is `r interval("early", "random walk", "nowcast", 11)`.

## Estimates six weeks after the last onset

```{r fit-late, results = "hide", message = FALSE, warning = FALSE}
late <- prepare(42)
fit_random_walk_late <- fit(late, random_walk)
estimates <- rbind(
  estimates,
  data.table(cutoff = "late", model_name = "random walk",
    weekly_estimates(fit_random_walk_late))
)
```

```{r late, fig.width = 7, fig.height = 4.5, fig.alt = "Weekly CFR estimates from the random-walk model six weeks after the last onset. The predicted fatal fraction equals the realised fraction in each week with almost no uncertainty, while prob keeps its intervals."}
plot_weekly("late", late)
```

Five more weeks of follow-up give nearly every fatal case time to die and make the survivors near-certain.
The predicted fatal fraction is now close to the realised fraction in every week, with almost no uncertainty.
The naive ratio has caught up with it.
`prob` keeps an interval, because it estimates the underlying CFR.
More follow-up has narrowed it, to `r interval("late", "random walk", "prob", 11)` for week 11.

## Other settings

In other settings only the data preparation and the `fit_cfr()` call change.
The code for the weekly estimates stays the same, because `posterior_prob_death()` reads the fit's own settings:

- With recorded recoveries (a `recovery_date` column and a `recovery_delay`), an unresolved case's probability is updated with both delays.
  It moves towards 0 or 1 sooner, which makes the recent weeks lean less on `prob`.
- With a `loss_prior`, a case followed well past both delays settles at a value set by each outcome's chance of being lost.
  Its outcome stays uncertain.
- With per-case cut-offs or a `last_contact_date`, each case is updated with its own follow-up.

## Interpreting the two estimates

`prob` is the model's CFR for a case from that week.
With a constant formula it is one number for the whole outbreak.
With a random walk, each week's value rests on that week's resolved cases and on its neighbours.
For the latest weeks, which have few resolved cases, it mostly follows the week before.

The predicted fatal fraction estimates the fraction of that week's cases that die.
For a recorded death it uses what happened.
For a case still unresolved it uses the model's `prob`, lowered by the time passed since symptom onset.
Its interval covers the model's parameters and whether each unresolved case dies, but it assumes the model is right.
For the latest weeks it is only as good as the model's `prob` for those weeks.
Once the fatal cases have died and the survivors have been followed well past the delay, the interval is close to zero width, even though the underlying CFR is still uncertain.
